%% acknowledgments.tex

% From mitthesis package
% Version: 1.02, 2024/06/19
% Documentation: https://ctan.org/pkg/mitthesis

\chapter*{Acknowledgments}
\pdfbookmark[0]{Acknowledgments}{acknowledgments}

I am grateful for Professor Aram Harrow who was both my instructor for the courses 8.370 and 8.372 at MIT and my thesis supervisor throughout my time as a Master of Engineering student. I sincerely appreciate the opportunity he provided to work as a research assistant in the Center for Theoretical Physics at the frontier of quantum computational chemistry, one of the most promising areas of quantum computing. Also crucial to this project was visiting PhD student Jakob Guenther from the University of Copenhagen. I appreciate his generosity with his time to have recurring meetings to transmit to me his expertise on this specific problem both in person and across a six-hour time difference. I also congratulate him on his upcoming post-doctorate position at MIT.

I also want to thank Professor Marek Perkowski of Portland State University, who taught me quantum computing before I came to MIT. Without his instruction and the work that we did together, I could not have arrived at this point in my academic career.

This material is based upon work supported by the U.S. Department of Energy, Office of Science, National Quantum Information Science Research Centers, and the Quantum Systems Accelerator. Funding was also provided by the Wellcome Leap as part of the Quantum for Bio (Q4Bio) Program.